---
title: "Phenology Flagging"
author: "Megan Vahsen"
date: "2023-04-17"
output:
  html_document:
    toc: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description of flags

This document details the code that informs how phenology data from the Bromecast 2022 experiments are flagged. The functions that are used to flag the phenology data are in the file "QAQC_functions.R" and are described in detail below.

```{r datamanip, echo = F, message = F}
# Load libraries
library(tidyverse)
library(here)
library(patchwork)
library(lme4)
library(mgcv); library(gratia)
library(geomtextpath)

# Read in derived phenology data
phen <- read_csv(here("gardens/deriveddata/SS2022_growthphenology_by_plantID.csv"))
# Read in plant ID info
ids <- read_csv(here("gardens/deriveddata/SS2022_plantID.csv"))
# Read in garden treatment info
gardens <-read_csv(here("gardens/rawdata/garden_treatments.csv"))
# Read in flagging data
flags <- read_csv(here("gardens/deriveddata/SS2022_flags.csv"))

# Merge together datasets
phen_id <- merge(phen, ids)

# Rename 'garden' column and remove cum_plot column
gardens %>% 
  mutate(site = garden) %>% 
  select(-cum_plot, -garden) -> gardens_sub

# Merge together datasets
phen_id_garden <- merge(phen_id, gardens_sub)

# Bring in flag data
flags <- read_csv(here("gardens/deriveddata/SS2022_flags.csv"))

# Merge together datasets
phen <- merge(phen_id_garden, flags)

# Set appropriate factors for variables
phen %>% 
  mutate(block = as.factor(block),
         plot = as.factor(plot),
         growout = as.factor(growout),
         density = as.factor(density),
         gravel = as.factor(gravel),
         genotype = as.factor(genotype)) %>% 
  mutate(plot_unique = as.factor(paste(block, plot, sep = "_")))-> phen
```


### Late emergence
Notes if the plant emerged after the first date plants were assessed for emergence, and if so, records the date of the visit where emergence was first observed. NAs for plants with recorded emergence on the first visit and for plants that never emerged.
```{r flag_lateemergence, eval = F}
flag_lateemergence <- function(plant_data){
  # was plant absent first visit, later present?
  if(plant_data$live[1]=="Y"){
    return(NA)  # emergence on first visit
  }else if(sum(plant_data$live=="Y")==0){
    return(NA)  # never emerged
  }else{
    # return date of first recorded presence
    tmp <- which(plant_data$live=="Y")
    if(length(tmp)>1) tmp <- tmp[1] 
    return(plant_data$jday[tmp])
  }
}
```

### Emergence date
Records emergence date. NAs for plants that never emerged.

```{r get_emergence_date, eval = F}
get_emergence_date <- function(plant_data){
  # date of first presence
  if(sum(plant_data$live=="Y")==0){
    return(NA)  # never emerged
  }else{
    # return date of first presence
    tmp <- which(plant_data$live=="Y")
    if(length(tmp)>1) tmp <- tmp[1] 
    return(plant_data$jday[tmp])
  }
}
```

```{r emergence, echo = F, warning=F, fig.width=6, fig.height=4, fig.align='center'}
# 1. Late emergence / emergence date
phen %>% 
  select(plantID, emergence_date) %>% 
  distinct() %>% 
  group_by(emergence_date) %>% 
  summarize(n = n()) -> emergence_summary

# Calculate proportion of NAs
emergence_summary %>% 
  mutate(is_complete = ifelse(is.na(emergence_date), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> emergence_NAs

# Create plot of data (remove NAs)
emergence_summary %>% 
  ggplot(aes(x = emergence_date, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("emergence date") +
  annotate("text", x = Inf, y = Inf, label = paste(emergence_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> emergence

emergence
```

### Death date
Records first date for which there absence recorded following a presence being recorded. Returns NAs if plants were alive and never died or if plants never emerged to begin with.
```{r, eval = F}
get_death_date <- function(plant_data){
  # date of first absence after a presence
  out <- NA  # death not observed
  tmp <- ifelse(plant_data$live=="Y",1,0) 
  tmp <- diff(tmp)
  deaths <- which(tmp==-1)
  if(length(deaths)>0){
      out <- plant_data$jday[1+min(deaths)]
  }
  return(out)
}
```

```{r death, fig.width=6, fig.height=4, fig.align='center', echo = F, warning = F}
# Get summary stats
phen %>% 
  select(plantID, death_date) %>% 
  distinct() %>%
  group_by(death_date) %>% 
  summarize(n = n()) -> death_summary

# Calculate proportion of NAs
death_summary %>% 
  mutate(is_complete = ifelse(is.na(death_date), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> death_NAs

# Create plot of data (remove NAs)
death_summary %>% 
  ggplot(aes(x = death_date, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("death date") +
  annotate("text", x = Inf, y = Inf, label = paste(death_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> death

death

```

### Resurrection
Records if plant was recorded as being present after previously being recorded as absent. Returns NAs if plants were never recorded as absent after being present or if plants never emerged to being with.

```{r, flag_resurrection, eval = F}
flag_resurrection <- function(plant_data){
  # was plant observed alive after being observed dead?
  out <- NA
  tmp <- ifelse(plant_data$live=="Y",1,0) 
  tmp <- diff(tmp)
  births <- which(tmp==1)
  deaths <- which(tmp==-1)
  if(length(births)>0 & length(deaths)>0){
    if(max(births)>min(deaths)){
      out <- plant_data$jday[1+max(births)]
    }
  }
  return(out)
}
```

```{r resurrection, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
# Get summary stats
phen %>% 
  select(plantID, resurrection_date) %>% 
  distinct() %>%
  group_by(resurrection_date) %>% 
  summarize(n = n()) -> resurrection_summary

# Calculate proportion of NAs
resurrection_summary %>% 
  mutate(is_complete = ifelse(is.na(resurrection_date), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> resurrection_NAs

# Create plot of data (remove NAs)
resurrection_summary %>% 
  ggplot(aes(x = resurrection_date, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("resurrection date") +
  annotate("text", x = Inf, y = Inf, label = paste(resurrection_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> resurrection

resurrection
```

### Phenological regression
Determines if phenology goes backwards through time (*e.g.*, V1 recorded after a previously recorded V2) and if so, records the maximum number of stages of regression for each plant. Returns NAs if phenological regression doesn't happen or if the plant never emerged to being with.

```{r flag_phenoregress, eval = F}
flag_phenoregress <- function(plant_data,pheno_table){
  # does phenology go backwards? returns how # of stages of regression
  out <- NA
  if(sum(plant_data$live=="Y")>1){
    plant_data <- merge(plant_data,pheno_table)
    plant_data <- plant_data[order(plant_data$jday),]
    tmp <- diff(plant_data$v_numeric)
    if(min(tmp)<0) out <- min(tmp)
  }
  return(out)
}
```

```{r phenoregress, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
phen %>% select(plantID, pheno_regress) %>% 
  distinct() %>%
  group_by(pheno_regress) %>% 
  summarize(n = n()) -> phenoregress_summary

# Calculate proportion of NAs
phenoregress_summary %>% 
  mutate(is_complete = ifelse(is.na(pheno_regress), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> phenoregress_NAs

# Create plot of data (remove NAs)
phenoregress_summary %>% 
  ggplot(aes(x = pheno_regress, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("phenological regression max. difference (# stages)") +
  annotate("text", x = Inf, y = Inf, label = paste(phenoregress_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> phenoregress

phenoregress
```

### Growth regression
Determines if the leaf length of the plant goes backwards through time and if so, records the maximum amount of shrinkage in millimeters. Records NAs if there is no growth regression or if plant did not emerge to begin with.
```{r flag_growthregress, eval = F}
flag_growthregress <- function(plant_data){
  # does leaf length go backwards? return max mm of shrinkage
  out <- NA
  if(sum(!is.na(plant_data$length_mm)>1)){
    tmp <- diff(plant_data$length_mm)
    if(sum(!is.na(tmp)>0)){
       if(min(tmp,na.rm=T)<0) out <- min(tmp,na.rm=T)
    }
  }
  return(out)
}
```

```{r growthregress, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
phen %>% select(plantID, growth_regress_mm) %>% 
  distinct() %>% 
  group_by(growth_regress_mm) %>% 
  summarize(n = n()) -> growthregress_summary

# Calculate proportion of NAs
growthregress_summary %>% 
  mutate(is_complete = ifelse(is.na(growth_regress_mm), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> growthregress_NAs

# Create plot of data (remove NAs)
growthregress_summary %>% 
  ggplot(aes(x = growth_regress_mm, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("growth regression max. difference (mm)") +
  annotate("text", x = Inf, y = Inf, label = paste(growthregress_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> growthregress

growthregress
```

### Frost heave
Determines if there was frost heave recorded for the plant and if so, returns the date of first recorded frost heave. Returns NAs if no frost heave recorded for the plant or if the plant never emerged to begin with. Also adds notes to observation about frost heave.

```{r flag_frostheave, eval = F}
flag_frostheave <- function(plant_data){
  out <- NA
  # date of first frost heave
  tmp <- which(plant_data$frost_heave=="Y")
  # add notes about frostheave
  tmp <- c(tmp,which(plant_data$standard_note=="frostheave"))
  # return date 
  if(length(tmp)>0){
    tmp <- min(tmp)
    out <- plant_data$jday[tmp]
  }
  return(out)
}
```

```{r frostheave, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
# Get summary stats
phen %>% 
  select(plantID, frostheave_date) %>% 
  distinct() %>%
  group_by(frostheave_date) %>% 
  summarize(n = n()) -> frostheave_summary

# Calculate proportion of NAs
frostheave_summary %>% 
  mutate(is_complete = ifelse(is.na(frostheave_date), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> frostheave_NAs

# Create plot of data (remove NAs)
frostheave_summary %>% 
  ggplot(aes(x = frostheave_date, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("frost heave date") +
  annotate("text", x = Inf, y = Inf, label = paste(frostheave_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> frostheave

frostheave
```

### Herbivory
Determines if there was herbivory recorded for the plant and if so, returns the date for which herbivory was first recorded. Returns NAs if there was no herbivory recorded or if plants never emerged to begin with. Also adds notes to observation about herbivory.
```{r, flag_herbivory, eval = F}
flag_herbivory <- function(plant_data){
  out <- NA
  # date of first herbivory observation
  tmp <- which(plant_data$herbivory=="Y")
  # add notes about herbivory
  tmp <- c(tmp,which(plant_data$standard_note=="herbivory"))
  # return date 
  if(length(tmp)>0){
    tmp <- min(tmp)
    out <- plant_data$jday[tmp]
  }
  return(out)
}
```

```{r herbivory, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
# Get summary stats
phen %>% 
  select(plantID, herbivory_date) %>% 
  distinct() %>%
  group_by(herbivory_date) %>% 
  summarize(n = n()) -> herbivory_summary

# Calculate proportion of NAs
herbivory_summary %>% 
  mutate(is_complete = ifelse(is.na(herbivory_date), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> herbivory_NAs

# Create plot of data (remove NAs)
herbivory_summary %>% 
  ggplot(aes(x = herbivory_date, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("herbivory date") +
  annotate("text", x = Inf, y = Inf, label = paste(herbivory_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> herbivory

herbivory
```


### Bad position
Determines if the plant was not in the expected position and if so, returns the date for which "bad position" was first recorded. Returns NAs if there was no "bad position: recorded or if plants never emerged to begin with. Also adds notes to observation about bad positioning.
```{r flag_badposition, eval = F}
flag_badposition <- function(plant_data){
  out <- NA
  # date of bad position observation
  tmp <- which(plant_data$standard_note=="badposition")
  if(length(tmp)>0){
    tmp <- min(tmp)
    out <- plant_data$jday[tmp]
  }
  return(out)
}
```

```{r badposition, fig.width=6, fig.height=4, fig.align='center', warning = F, echo = F}
# Get summary stats
phen %>% 
  select(plantID, bad_position) %>% 
  distinct() %>%
  group_by(bad_position) %>% 
  summarize(n = n()) -> bad_position_summary

# Calculate proportion of NAs
bad_position_summary %>% 
  mutate(is_complete = ifelse(is.na(bad_position), 0, 1)) %>% 
  group_by(is_complete) %>% 
  summarize(n = sum(n)) %>% 
  mutate(prop = n * 100 / sum(n)) %>% 
  filter(is_complete == 0) %>% 
  pull(prop) %>% round(2) -> bad_position_NAs

# Create plot of data (remove NAs)
bad_position_summary %>% 
  ggplot(aes(x = bad_position, y = n)) +
  geom_bar(stat="identity", width = 1, fill = "gray47") +
  geom_rug(sides = "b") +
  theme_classic(base_size = 14) +
  xlab("bad position date") +
  annotate("text", x = Inf, y = Inf, label = paste(bad_position_NAs, "% NAs", sep = ""),
           vjust = 2, hjust = 1.1, size = 8) -> bad_position

bad_position
```

### Other notes
Determines if there are other standardized notes that are important to add to the derived data that are not herbivory, frost heave, or bad positioning, and if so, records them. If there are multiple notes, these are combined together. Returns NAs if there are no other notes.

```{r flag_other, eval = F}
flag_other <- function(plant_data){
  out <- NA
  # include other consequential notes
  tmp <- which(!is.na(plant_data$standard_note))
  if(length(tmp)>0){
    # remove frostheave, herbivory and badposition (these dealt with separately)
    tmp <- plant_data$standard_note[tmp]
    notes <- tmp[which(tmp != "herbivory" & tmp != "frostheave" & tmp != "badposition")]
    if(length(notes)>0){
      notes <- unique(notes)  # prevent duplicates
      notes <- paste(notes,collapse='-') # if multiple notes, combine
      out <- notes
    }
  }
  return(out)
}
```

```{r, echo = F}
# Print out proportion of plants with other notes
phen %>% 
  select(plantID, other) %>% 
  distinct() %>% 
  group_by(other) %>% 
  summarize(n = n()) %>% 
  mutate(percentage = round(n*100/sum(n),3))
```

## Running analyses to see the impact of flagging on inference

### Assessing survival rates
```{r, fig.height=6, fig.width = 12, fig.align='center', echo = F}
phen %>% 
  # Remove plants with bad position flags
  filter(is.na(bad_position),
         # Remove plants that were missing
         is.na(missing_plant)) %>% 
  # Create column to calculate difference between frost heave and emergence timing
  mutate(frost_time = frostheave_date - emergence_date) %>% 
  # Remove plants that had frost heave before emergence
  filter(frost_time > 0 | is.na(frost_time)) -> phen_sub1

# Create subsetted data that has one observation per plant
phen_sub1 %>% 
  group_by(plantID) %>% 
  slice(which.max(jday)) %>% 
  mutate(survived = ifelse(live == "Y", 1, 0)) %>%
  select(plantID, site, block, plot, plot_unique, frostheave_date, density, gravel, genotype, survived) %>% 
  distinct() -> phen_sub1_plant

# Fit emergence model
m1_gam <- gam(survived ~ density*gravel +
                # Random intercept for block
                s(block, bs = 're') + 
                # Random intercept for plot nested within block
                s(plot_unique, bs = 're') +
                # Random intercept for genotype
                s(genotype, bs = 're'),
                # Random slope for density x genotype effects
                #s(genotype, density, bs = 're')+
                # Random slope for gravel x genotype effects
                #s(genotype, gravel, bs = 're'),
              data = phen_sub1_plant, method = 'REML', family = "binomial")

# Get confidence intervals around variance components
variance_comp(m1_gam)

# Plot effects of genotype
confint(m1_gam, parm = "s(genotype)") %>% 
  ggplot(aes(x = reorder(genotype, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = genotype, xend = genotype)) +
  xlab("genotype") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  ylim(-1.5, 1.5) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> a

# Plot effects of block
confint(m1_gam, parm = "s(block)") %>% 
  ggplot(aes(x = reorder(block, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = block, xend = block)) +
  xlab("block")+
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  ylim(-1.5, 1.5)+
  theme_bw(base_size = 14)-> b

# Plot effects of unique plot
confint(m1_gam, parm = "s(plot_unique)") %>% 
  ggplot(aes(x = reorder(plot_unique, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = plot_unique, xend = plot_unique)) +
  xlab("plot") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  ylim(-1.5, 1.5)+
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> c

a / b / c
```

Variation in likelihood of survival is mostly driven by small, plot-level variation rather than large, block-level variation or genotypic variation.  

```{r, fig.height = 4, fig.width = 6, fig.align='center', echo = F}
# Get predicted means of density:gravel treatments
dg_means <- summary(emmeans::emmeans(m1_gam, ~density*gravel, non.nuis = c("density", "gravel"), type = "response"))

# Plot density:gravel effects
dg_means %>% 
  mutate(survived = prob) %>% 
  ggplot(aes(x = density, y = survived)) +
  geom_point() +
  geom_segment(aes(x = density, xend = density, y = lower.CL, yend = upper.CL)) +
  geom_jitter(data = phen_sub1_plant, aes(x = density, y = survived),
              height = 0.05, width = 0.3, alpha = 0.2, shape = 1) +
  facet_wrap(~gravel) +
  theme_bw(base_size = 14) +
  ylab("probability of survival")
```

The likelihood of survival is higher at high density than at low density, and this effect is consistent across gravel types. 

### Assessing flowering rates (i.e., if flowered or not, based on plants that survived)
```{r, fig.align='center', fig.width=12, fig.height=6, echo = F}
# Create column for if flowered during the course of the experiment. First get
# plants that flowered at least once during the experiment but also survived
phen_sub1_plant %>%
  filter(survived == 1) %>% 
  pull(plantID) -> survived_plants

phen_sub1 %>% 
  filter(v %in% c("FG") & plantID %in% survived_plants) %>% 
  group_by(plantID) %>%
  slice(which.min(jday)) %>% 
  ungroup() %>% 
  mutate(flowered = 1) -> flowered_plants

# Subset to get plants that did not flower
`%notin%` <- Negate(`%in%`) # Not in operator

phen_sub1 %>% 
  filter(plantID %in% survived_plants) %>% 
  filter(plantID %notin% flowered_plants$plantID) %>%
  group_by(plantID) %>% 
  slice(which.min(jday)) %>% 
  ungroup() %>% 
  mutate(flowered = 0) -> unflowered_plants

# Join together flowered and unflowered data sets
phen_sub1_plant3 <- rbind(flowered_plants, unflowered_plants)

# Fit flowering model
m3_gam <- gam(flowered ~ density*gravel +
                # Random intercept for block
                s(block, bs = 're') + 
                # Random intercept for plot nested within block
                s(plot_unique, bs = 're') +
                # Random intercept for genotype
                s(genotype, bs = 're'),
              data = phen_sub1_plant3, method = 'REML', family = "binomial")

# Get confidence intervals around variance components
variance_comp(m3_gam)

# Plot effects of genotype
confint(m3_gam, parm = "s(genotype)") %>% 
  ggplot(aes(x = reorder(genotype, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = genotype, xend = genotype)) +
  xlab("genotype") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> h

# Plot effects of block
confint(m3_gam, parm = "s(block)") %>% 
  ggplot(aes(x = reorder(block, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = block, xend = block)) +
  xlab("block")+
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14)-> i

# Plot effects of unique plot
confint(m3_gam, parm = "s(plot_unique)") %>% 
  ggplot(aes(x = reorder(plot_unique, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = plot_unique, xend = plot_unique)) +
  xlab("plot") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> j

h / i / j
  
```
Variation in likelihood of flower (given that a plant survived) is mostly driven by genotypic variation rather than plot- or block-level variation.   

```{r, fig.height = 4, fig.width = 6, fig.align='center', echo = F}
# Get predicted means of density:gravel treatments
dg_means_flower <- summary(emmeans::emmeans(m3_gam, ~density*gravel, non.nuis = c("density", "gravel"), type = "response"))

# Plot density:gravel effects
dg_means_flower %>% 
  mutate(flowered = prob) %>% 
  ggplot(aes(x = density, y = flowered)) +
  geom_point() +
  geom_segment(aes(x = density, xend = density, y = lower.CL, yend = upper.CL)) +
  #geom_jitter(data = phen_sub1_plant3, aes(x = density, y = flowered),
              #height = 0.05, width = 0.3, alpha = 0.1, shape = 1) +
  facet_wrap(~gravel) +
  theme_bw(base_size = 14) +
  ylab("probability of flowering (if survived)")
```

The probability of flowering is higher at high density than at low density and higher when the plant is in black gravel compared to white gravel.

## Assessing time to flower
```{r, fig.align='center', fig.width=12, fig.height=6, echo = F}
# Calculate when the earliest flowering day is for each plant. This procedure
# removes plants that do not flower of the course of the experiment.
phen_sub1 %>%
  filter(v == "FG") %>%
  group_by(plantID) %>%
  slice(which.min(jday)) -> phen_sub1_plant2

# Get summary statistics
# phen_sub1_plant2 %>% 
#   group_by(jday) %>% 
#   summarize(n = n())

# Try fitting an ordered categorical model for the response variable (because sampled every 2 weeks)
phen_sub1_plant2 %>% 
  mutate(ord = case_when(jday %in% 123:125 ~ 1,
                         jday %in% 138:140 ~ 2,
                         jday %in% 151:154 ~ 3,
                         jday %in% 165:166 ~ 4)) -> phen_sub1_plant2

phen_sub1_plant2 %>% 
  mutate(trt = factor(paste(density, gravel, sep = "_"))) -> phen_sub1_plant2

m2_gam <- gam(ord ~ trt +
              # Random intercept for block
                s(block, bs = 're') + 
                # Random intercept for plot nested within block
                s(plot_unique, bs = 're') +
                #s(genotype, bs = 're'),
                # Random slope for genotype x density:gravel?
                s(genotype, by = trt, bs = 're'),
              family = ocat(R = 4),
              data = phen_sub1_plant2)

# Plot effects of genotype
confint(m2_gam, parm = "s(genotype)") %>% 
  ggplot(aes(x = reorder(genotype, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = genotype, xend = genotype)) +
  xlab("genotype") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> d

# Plot effects of block
confint(m2_gam, parm = "s(block)") %>% 
  ggplot(aes(x = reorder(block, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = block, xend = block)) +
  xlab("block")+
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14)-> e

# Plot effects of unique plot
confint(m1_gam, parm = "s(plot_unique)") %>% 
  ggplot(aes(x = reorder(plot_unique, est), y = est)) +
  geom_point() +
  geom_segment(aes(y = lower, yend = upper, x = plot_unique, xend = plot_unique)) +
  xlab("plot") +
  ylab("estimate") +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))-> f

d / e / f


```

Variation in the timing of flowering is mostly driven by genotypic variation, but also by small, plot-level variation.

```{r, fig.align='center', fig.width=6, fig.height=5, echo = F}
new_data <- expand.grid(density = unique(phen_sub1_plant2$density),
                        plot_unique = unique(phen_sub1_plant2$plot_unique),
                        genotype = unique(phen_sub1_plant2$genotype),
                        gravel = unique(phen_sub1_plant2$gravel)) %>%
  mutate(block = as.factor(str_extract(plot_unique, ".+?(?=_)")))

# new_data <- expand.grid(trt = unique(phen_sub1_plant2$trt),
#                         plot_unique = unique(phen_sub1_plant2$plot_unique),
#                         genotype = unique(phen_sub1_plant2$genotype)) %>%
#   mutate(block = as.factor(str_extract(plot_unique, ".+?(?=_)")))

# Predict means across all possible combinations
predict(m2_gam, newdata = new_data, type = "response") -> m2_preds
# Group together with data
new_data %>% 
  mutate(prob1 = m2_preds[,1],
         prob2 = m2_preds[,2],
         prob3 = m2_preds[,3],
         prob4 = m2_preds[,4]) -> pred_data
# Average across combinations of interest
pred_data %>% 
  group_by(density, gravel) %>% 
  summarize(across(where(is.numeric), mean)) %>% 
  mutate(cum_prob_1 = prob1,
         cum_prob_12 = prob1 + prob2,
         cum_prob_123 = prob1 + prob2 + prob3,
         cum_prob_1234 = 1) %>% 
  gather(key = cat, value = cum_prob, cum_prob_1:cum_prob_1234) %>% 
  mutate(jday = case_when(cat == "cum_prob_1" ~ 124,
                          cat == "cum_prob_12" ~ 139,
                          cat == "cum_prob_123" ~ 152,
                          cat == "cum_prob_1234" ~ 165),
         treatment = paste(density, gravel, sep = "_"),
         density = ifelse(density == "hi", "high", "low")) %>% 
  ggplot(aes(x = jday, y = cum_prob, group = treatment, color = gravel, shape = density, linetype = density)) +
  geom_textline(aes(label = paste(density, gravel, sep = "-")), hjust = 0.42) +
  geom_point(size = 3) +
  ylab("cumulative probability of flowering by date") +
  theme_classic(base_size = 14) +
  xlab("julian day") +
  scale_color_manual(values = c("black", "gray70")) +
  scale_x_continuous(breaks = c(124,139,152,165)) +
  theme(legend.position = "none")

# pred_data %>% 
#   group_by(genotype, trt) %>% 
#   summarize(across(where(is.numeric), mean)) %>% 
#   mutate(cum_prob_1 = prob1,
#          cum_prob_12 = prob1 + prob2,
#          cum_prob_123 = prob1 + prob2 + prob3,
#          cum_prob_1234 = 1) %>% 
#   gather(key = cat, value = cum_prob, cum_prob_1:cum_prob_1234) %>% 
#   mutate(jday = case_when(cat == "cum_prob_1" ~ 124,
#                           cat == "cum_prob_12" ~ 139,
#                           cat == "cum_prob_123" ~ 152,
#                           cat == "cum_prob_1234" ~ 165)) %>%
#   filter(genotype %in% c("1", "2", "5", "6")) %>% 
#   ggplot(aes(x = jday, y = cum_prob, group = genotype:trt, color = genotype)) +
#   geom_line() +
#   ylab("cumulative probability of flowering by date") +
#   theme_classic(base_size = 14) +
#   xlab("julian day") +
#   scale_x_continuous(breaks = c(124,139,152,165))
  
```

Plants flower earlier when at a higher density than at a lower density, and flower earlier in black gravel than in white gravel. The effect of density on flowering time is greater when plants are in black gravel than when plants are in white gravel. 



